---
title: "Practical Machine Learning Project"
author: "Omero Catan"
date: "August 10, 2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Provided Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Project Goal

The goal of the project is to predict the manner in which the exercise was conducted. This is defined as the "classe" variable in the training set data. Once a model is built, it will be used to predict 20 different test cases.

## Load Required Packages and Data Sets
```{r}
library(caret)
library(randomForest)

trainingdat <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"))
testingdat <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"))

dim(trainingdat); dim(testingdat)
str(trainingdat)
```

## Data Set Specifics

As presented, both data sets contain 160 variables. The training set has 19,622 observations while the test set is the 20 observations referenced in the "Project Goal" section. When examining the data, you will observe that there are some variables that are mostly not applicable for the model. Additionally, the first 7 variables are identifiers/descriptors of when/how that particular observation was ran. These variables will not impact the effectiveness of our predictive model and can be removed from the data.

```{r}
delete_training <- which(colSums(is.na(trainingdat) | trainingdat == "") > 0.75 * dim(trainingdat)[1])
trainingdat <- trainingdat[, -delete_training]
trainingdat <- trainingdat[, -c(1:7)]

delete_testing <- which(colSums(is.na(testingdat) | testingdat == "") > 0.75 * dim(testingdat)[1])
testingdat <- testingdat[, -delete_testing]
testingdat <- testingdat[, -c(1:7)]

dim(trainingdat); dim(testingdat)
str(trainingdat)
```

## Clean Data

The clean data contains 53 of the original 160 variables. There are multiple methods to use when creating a predictive model. Three include CLassification Trees with K-Fold Cross Validation, Random Forests, and Boosting.

When using Classification Trees with K-Fold Cross Validation, the larger the number of folds ("K"), the less bias and more variance you will experience. With Random Forests, at each split you bootstrap the variables to grow multiple trees and vote. With Boosting, the goal is to calculate weights of variables based upon errors.

My original plan was to test the three methods above and select the method with the highest accuracy to run the test cases on. However, I kept comming across computer memory limitations.

In this project, we are attempting to accuratly predict the manner in which the exercise was conducted. Accuracy is a proficency of the Random Forests method so I selected that method to base the model on.

## Random Forests
```{r}
#Split training data into training and tests sets.
set.seed(0911)
inTrain <- createDataPartition(trainingdat$classe, p = 0.7, list = FALSE)
training <- trainingdat[inTrain, ]
testing <- trainingdat[-inTrain, ]

dim(training); dim(testing)

modFit <- randomForest(classe ~., data = training)
predictFit <- predict(modFit, testing, type = "class")
confuse <- confusionMatrix(predictFit, testing$classe)
confuse
```

## Random Forests Output

In random forests, the unbiased estimate of the test set error is estimated internally during the run of the model. Therefore, cross-validation is not required to be ran separately.

The model created has an accuracy of 99.41%. Therefore, my predcted accuracy for the out-of-sample error is 0.59%. This is an acceptable accuracy percentage to move forward and apply to the test cases.

## Test Cases
```{r}
test_cases <- predict(modFit, testingdat, type = "class")
test_cases
```